\section{Sequences: A Brief Introduction}

Generally speaking, we can think of a (real) sequence as being an infinite list of real numbers that can be indexed by the natural numbers. That is, a sequence has a well-defined first element, second element, third element, and so on.

It is important to note that certain things cannot be ``sequenced": for instance, a famous argument known as \textbf{Cantor's Diagonal Argument} shows that you cannot fit every single real number into a sequence. Indeed, it isn't too difficult to see intuitively that if you look at the interval $\brac{0,1}$, you cannot possibly list down every single element. You can think of this as being due to the fact that you can have as many decimal places as you want, for example.

Anyway, the bottom line is, a sequence is a list of numbers as well as a well-defined way of enumerating them. We usually denote sequences by something like
\begin{align*}
    \parenth{a_n}_{n=1}^{\infty}
\end{align*}
consisting of elements $a_1, a_2, a_3, a_4, \ldots$ which all stand for real numbers. The letter $a$ is used to identify the sequence---that is, it is the ``name" of the sequence---and the subscripts, also called indices, are used to denote various numbers in the sequence. Formally speaking, one can think of a sequence as a function taking in natural numbers (ie, some element of $\N = \set{1, 2, 3, \cdots}$) and outputting real numbers. So, for every index $n$, there is some real number $a_n$ associated with the sequence $a$.

\input{AS}
\input{GS}
% \input{IS}  % Could potentially add a section on infinite series in the future...